---
title: "Lecture8 Diagnostics"
author: "Witold Wolski"
date: "March 29, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(lme4)
library(ggplot2)
```

Start with a plot of the data, to look for outliers
```{r}

load('MAS473.RData')
attach(ergoStool)
qplot(Type, effort, facets=~Subject,data=ergoStool)

fm1<-lmer(effort~Type -1 + (1|Subject),ergoStool)
```

# 3.5.6 assessing the fitted model

Expected values of random effects
```{r}
ranef(fm1)

plot(ranef(fm1)$Subject, pch=16)
```
# Check assumption that random effects are normally distributed
```{r}
qqnorm(unlist(ranef(fm1)))
abline(0,1.332) # Reference gradient is estimated standard deviation of random effect

```
For background, R uses these quantiles for the QQ plot
Do ?ppoints for more details
```{r}
(qu<-(1:9 - 3/8)/(9 + (1-3/8)-3/8))
z<-qnorm(qu)
plot(z,sort(unlist(ranef(fm1)$Subject)) )
```
Fitted values
```{r}
fitted(fm1)
```

Residuals (not standardaised)

```{r}
resid(fm1)
```

Plot level 1 fitted values against residuals

```{r}

qplot(fitted(fm1),resid(fm1), data = ergoStool)
qplot(fitted(fm1), resid(fm1), facets=~Type, data = ergoStool) # Residual plots by stool type
qplot(fitted(fm1), resid(fm1), facets=~Subject, data = ergoStool) # Residual plots by subject

```

Check assumption that errors are normally distributed
```{r}
qqnorm(resid(fm1))
#qqline(resid(fm1))
abline(0,1.1003)
```

Plot level 0 fitted values against residuals
```{r}
(fitted.level0<-fm1@pp$X %*% fixef(fm1))
resid.level0<-effort-fitted.level0
qplot(fitted.level0,resid.level0)

```

Can jitter the points to make easier to see
```{r}
qplot(fitted.level0,resid.level0)+geom_jitter() 
```


Assess general fit of model

```{r}
plot(fitted(fm1),ergoStool$effort)
abline(0,1)
```

# Section 4.1.1 Oxides example

Plot the data first to check for outlying lots, wafers within lots, sites within wafers


```{r}
attach(Oxide)
qplot(Wafer, Thickness, facets=~Lot, data=Oxide)
```

```{r}
fm1<-lmer(Thickness~1+(1|Lot/Wafer),data=Oxide)
summary(fm1)
```

Estimated random effects
```{r}
ranef(fm1)
```

# Check model assumptions
```{r}
qqnorm(unlist(ranef(fm1)$`Wafer:Lot`),ylab="Wafer-within-lot-level random effects")
abline(0,5.9891)
qqline(unlist(ranef(fm1)$`Wafer:Lot`))

qqnorm(unlist(ranef(fm1)$Lot),ylab="Lot-level random effects")
abline(0,11.3967)
```

Doesn't look so good, but sample size is small

```{r}
qqnorm(resid(fm1))
abline(0, 3.5453)
qqline(resid(fm1))

plot(fitted(fm1),residuals(fm1))
```

```{r}
qplot(fitted(fm1), residuals(fm1), facets=~Lot, data=Oxide)

```
Why isn't the following plot helpful?
```{r}
qplot(fitted(fm1),residuals(fm1), facets=~Wafer, data=Oxide)
```

Assess overall model fit
```{r}
plot(fitted(fm1),Thickness)
abline(0,1)
```

Three levels of fitted values
In matrix notation (Section 2.6), can extract Z with

```{r}
(Zt<-fm1@pp$Zt) # this is transpose of Z
```

Convert into normal matrix form, and take transpose
```{r}
Z<-as.matrix(t(Zt))
```

Put the predicted random effects into single column vector
```{r}
b.hat<-as.matrix(rbind(ranef(fm1)$`Wafer:Lot`,ranef(fm1)$Lot ))
```

Inner most level is 'level 2'


$$
\hat{y_{ijk}} = \hat{beta} + \hat{b_i} + \hat{b_{ij}}
$$

Calculate manually

```{r}
fitted.level2<-as.matrix(fm1@pp$X %*% fixef(fm1) + Z%*%b.hat)
```
inner most level also given by the fitted command

```{r}
fitted(fm1)
```

Check they are the same

```{r}
cbind(fitted.level2, fitted(fm1))
```

Level 1

$$
\hat{y_{ijk}} = \hat{beta} + \hat{b_i} 
$$

```{r}
fitted.level1<-as.matrix(fm1@pp$X %*% fixef(fm1) + Z[,25:32]%*%b.hat[25:32,1])
resid.level1<-Thickness-fitted.level1
plot(fitted.level1, resid.level1)
```

Level 0 (not very informative, as only a single fixed effect parameter)

$$
\hat{y_{ijk}} = \hat{beta} 
$$

```{r}
(fitted.level0<-as.matrix(fm1@pp$X %*% fixef(fm1) ))
```

