---
title: "Oats"
author: "Dr Richard Wilkinson"
date: "8 February 2016"

output:
  pdf_document: default
  html_document: default
---
# jsp data

```{r}
library(faraway)
library(lme4)
library(ggplot2)
data(jsp)
head(jsp)
dim(jsp)
```

Select final year only

```{r}
jspr <- jsp[jsp$year==2,]
qplot(social, math, data = jspr, geom='boxplot')
qplot(school, math, data = jspr, geom='boxplot')
```

Lets account for school

```{r}
lmer(math~ social + (1|school), data=jspr, contrasts=list(social=contr.sum))
```

To account for class we need to nest effects

```{r}
lmer(math~ social + (1|school)+(1| school:class), data=jspr, contrasts=list(social=contr.sum))

```

```{r}
lmer(math~ social + (1| school/class), data=jspr, contrasts=list(social=contr.sum))

```


# Chapter 3 Section 3.1.1
```{r}
load('MAS473.RData')
```

## Inspect the data
```{r}
head(Oxide)
```

Note that Wafer isn't a 3-level factor; Wafer is nested within Lot
eg $Y_{121}$ and $Y_{221}$ do not refer to the same Wafer.

```{r}
library(ggplot2)
qplot(data=Oxide, Wafer, Thickness, facets=~Lot)
qplot(data=Oxide, Wafer, Thickness, facets=~Lot, geom='boxplot')
```

which do you prefer?

```{r}
qplot(data=Oxide,Lot,Thickness,geom='boxplot')
```

We can see variation between Wafers in the same Lot
and variation between Lots

```{r}
fm1<-lmer(Thickness~1+(1|Lot/Wafer),data=Oxide)
summary(fm1)

```

equivalent command
```{r}
fm2 <- lmer(Thickness~1 +(1|Lot)+(1|Lot:Wafer), data =Oxide)
summary(fm2)
# Estimated random effects
ranef(fm1)

```


# Section 3.1.2

Split plot example

```{r}

attach(Oats)
qplot(Block, yield, geom='boxplot')
qplot(nitro, yield)
qplot(Variety,yield, geom='boxplot')

qplot(Variety:Block, yield, geom='boxplot', col=Variety)
```

Can get the plot random effect via Block/Variety
Variety included as a fixed effect too

```{r}
(fm1<-lmer(yield~ nitro + Variety + (1|Block/Variety),Oats))
```
doesn't mean that variety is a random effect, just that it indicates the different plots within the block
note that we don't need a random effect for the subplot - only 1 observation in each so represented by 

$$\epsilon_{ijk}$$

If this confuses you, can specify a plot factor:

```{r}
Plt<-gl(18,4)
Oats<-data.frame(Oats,Plt)
(Oats)
```

Same model as fm1
```{r}
(fm1b<-lmer(yield~nitro+Variety+(1|Block)+(1|Plt),Oats))

```

Model is

$$Y_{ijk} = \mu + \tau_{v(i,j)} + \beta x_{ijk} + b_i + b_{ij} + \epsilon_{ijk}$$
where

$$ 
\begin{aligned}
i&=1,...,6: \textrm{block}\\
j&=1,2,3: \textrm{plot} \\
k&=1,2,3,4: \textrm{subplot} \\
v(i,j) &= 1,2,3 ~\textrm{corresponding to variety  (eg. v(1,2)=2 for "Golden Rain")}\\
x_{ijk} & ~\textrm{nitro level}
\end{aligned}
$$

```{r}
summary(fm1b)
ranef(fm1b)
```
Note: can't have an ordinary linear model with plot effect, as confounded with Variety

```{r}

(lm1<-lm(yield~nitro+Variety+Plt ,Oats, contrasts=list(Plt=contr.sum)))
```

The 'equivalent' fixed effects model can be achieved by having a Variety:Block interaction
Gives same estimates for Variety effects, but with smaller standard errors.


```{r}
lm1<-lm(yield~nitro+Variety*Block ,Oats, contrasts=list(Block=contr.sum))
summary(lm1)
```
