---
title: "Lecture 1 - fixed vs random effects models"
author: "Dr Richard Wilkinson"
date: "8 February 2016"
output: html_document
---

```{r}
library(lme4)
library(lattice)
load('MAS473.RData')
attach(ergoStool)

```

# Section 3.5.2 

## inspecting the data

```{r}
str(ergoStool)
head(ergoStool)
tail(ergoStool)
xtabs(~ Type + Subject, ergoStool)
matrix(by(effort,list(Type,Subject),mean),4,9)
```

## Plotting the data

# With lattice
```{r}
xyplot(effort~Type|Subject,data=ergoStool)
```

# With ggplot2

```{r}
library(ggplot2)
qplot(Type, effort, facets=~Subject, data =ergoStool)
```

```{r, include =FALSE}
dotplot(reorder(Subject,effort) ~ effort, data=ergoStool, groups = Type, type = c("p"),pch=Type, par.settings=simpleTheme(pch=Type), xlab = "Effort to arise (Borg scale)", auto.key = list(columns=4))
```

```{r, include=FALSE}
qplot(reorder(effort, effort), reorder(Subject,effort),  pch=Type, data =ergoStool, color=Type,xlab = "Effort to arise (Borg scale)")
```

the argument reorder(Subject,effort) arranges the groups ("Subject") in order of increasing mean effort. 

## Boxplots

```{r}
plot(Type,effort, xlab="Type", ylab="effort")
```

```{r}
qplot(Type, effort, geom='boxplot')
```


```{r}
par(mfrow=c(1,2))
plot(Subject,effort, xlab="Subject", ylab="effort")
plot(reorder(Subject,effort),effort, xlab="Subject", ylab="effort") 
```

```{r}
qplot(reorder(Subject, effort), effort, geom='boxplot', xlab='Subject')
```

```{r}
plot.design(ergoStool)
```

# 3.5.3 Fitting the model

## Cell means formulation
```{r}
(fm1<-lmer(effort~Type-1 + (1|Subject),ergoStool))
```


## Default formulation
```{r}
(fm2<-lmer(effort~Type + (1|Subject),ergoStool))
```

## Sum to zero formulation

```{r}
options(contrasts = rep("contr.sum", 2))
(fm3<-lmer(effort~Type + (1|Subject),ergoStool))#, contrast = ))
```

go back to default treatment constrasts.

```{r}
options (contrasts = rep("contr.treatment", 2))
```

Same but without using side effects

```{r}
(fm4<-lmer(effort~Type + (1|Subject),ergoStool, contrast = list(Type = contr.sum)))

```

```{r}
summary(fm1)

```

# Compare fixed effects estimates with
```{r}
by(effort,Type,mean)
mean(effort[Type=='T1'])
```

Standard errors for the fixed effects estimates
$(1.7755/9+1.2106/9)^{0.5}$


Correlation between fixed effects estimates
$1.7755/(1.7755+1.2106)$


Compare fixed and random effects


```{r}
fm1
(lm1<-lm(effort~Type-1+Subject,ergoStool, contrasts = list(Subject=contr.sum)))
```

```{r}
summary(lm1)
summary(fm1)

```

Note the standard errors are smaller.
Why?


